---
title: "Wikipedia Portal Traffic and JavaScript Support"
author: "Mikhail Popov (Analysis & Report), Julien Girault (UX Engineering)"
date: "February 10, 2016"
output: 
  pdf_document: 
    fig_height: 6
    fig_width: 10
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
library(magrittr)
import::from(dplyr, group_by, summarize, mutate, select, rename, arrange, keep_where = filter, ungroup, top_n)
library(ggplot2)
correct_proportion <- function(x) {
  vapply(x, function(x, y) { return(min(x, y)) }, 0.0, y = 1)
}
```

```{r data}
aggregates <- readr::read_csv('portal-js.csv')
requests <- keep_where(aggregates, date > as.Date("2016-02-05"))
```

## Daily Downloads of Wordmark and JS File

To identify what proportion of Wikipedia Portal requests lack JavaScript (JS) support ([T124655](https://phabricator.wikimedia.org/T124655)), we decided to compare the number of sampled requests to the recently deployed JS file (see [T124713](https://phabricator.wikimedia.org/T124713)) relative to the number of sampled requests to the Wikipedia Wordmark. The rationale for this being that if a user visits wikipedia.org, their browser may or may not make a request for the JS file depending on their settings and JS support, but their browser will definitely make a request for the JS file. Note: some countries made more requests for the JS file than the wordmark image, which suggests that certain numbers and percentages should be taken with a grain of salt, and should be regarded as rough estimates.

Also, while we filtered out as much bot and spider traffic as we could prior to aggregation and analysis, but we would like the reader to note that our databases of bots' user agents (UAs) are not complete.

```{r daily_downloads, fig.width = 9, fig.height = 5.5}
aggregates %>%
  group_by(date) %>%
  summarize(index_js = sum(index_js), wikipedia_wordmark = sum(wikipedia_wordmark)) %>%
  ggplot(data = ., aes(x = date, y = wikipedia_wordmark)) +
  geom_line(color = "black", size = 1.1) +
  geom_point(color = "black", size = 3) +
  geom_line(aes(y = index_js), color = "cornflowerblue", size = 1.1) +
  geom_point(aes(y = index_js), color = "cornflowerblue", size = 3) +
  geom_text(aes(y = index_js + 6e4, label = sprintf("%.2f%%", 100*index_js/wikipedia_wordmark))) +
  scale_y_continuous(name = "Requests", breaks = seq(5e5, 3e6, 5e5), labels = polloi::compress, limits = c(0, 3e6)) +
  ggtitle("Daily JS support from sampled requests on Wikipedia Portal") +
  annotate("text", x = as.Date("2016-02-05"), y = 5e5, label = "JS patch deployed") +
  ggthemes::theme_gdocs() +
  theme(rect = element_blank()) +
  xlab("Date")
```

**Figure 1** suggests that, overall, 93% of Wikipedia Portal requests support JS. In this report, we break that 93% down by countries and browsers. The following **Table 1** summarizes JS support across 235 countries:

```{r js_support_table}
requests %>%
  group_by(date, country) %>%
  summarize(index_js = sum(index_js), wikipedia_wordmark = sum(wikipedia_wordmark)) %>%
  mutate(proportion = correct_proportion(index_js / wikipedia_wordmark)) %>%
  group_by(country) %>%
  summarize(proportion = median(proportion)) %>%
  summarize(`Min %` = min(proportion),
            `25th Percentile` = quantile(proportion, 0.25),
            `Average %` = mean(proportion),
            `Median %` = median(proportion),
            `75th Percentile` = quantile(proportion, 0.75),
            `Max %` = max(proportion)) %>%
  { .[1, ] <- sprintf("%.2f%%", 100 * .[1, ]); . } %>%
  knitr::kable(align = "r")
```

The median (88.31%) different than Fig. 1's 93% because in the figure we used the total of all requests, while the numbers in Tbl. 1 reflect a distribution of JS support in requests broken up by countries.

\newpage

```{r maps_setup, message = FALSE}
library(maps)
world_map <- map_data("world")
world_map$region[world_map$region == "USA"] <- "United States"
world_map$region[world_map$region == "UK"] <- "United Kingdom"
```

```{r js_support_choropleth}
worldwide_js_support <- requests %>%
  group_by(date, country) %>%
  summarize(index_js = sum(index_js), wikipedia_wordmark = sum(wikipedia_wordmark)) %>%
  keep_where(wikipedia_wordmark > 1e2) %>%
  mutate(proportion = correct_proportion(index_js / wikipedia_wordmark)) %>%
  group_by(country) %>%
  summarize(proportion = median(proportion))

ggplot(data = worldwide_js_support) +
  geom_map(data = world_map, aes(map_id = region), fill = "white", map = world_map, color = "black", size = 0.25) +
  geom_map(aes(map_id = country, fill = proportion), map = world_map) +
  expand_limits(x = world_map$long, y = world_map$lat) +
  scale_fill_gradientn(name = "% of requests with JS support: ",
                       limits = c(0.5, 1),
                       colors = RColorBrewer::brewer.pal(5, "YlGnBu"),
                       labels = scales::percent_format()) +
  ggtitle("JS support (from countries with >100 WP Wordmark requests)") +
  ggthemes::theme_fivethirtyeight() +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(),
        legend.key.width = unit(0.6, "in"))
```

For the *choropleth* -- a map shaded according to a statistic -- above, we excluded 199 countries with less than 100 sampled Wordmark requests to make the proportion estimates more reliable. For example, if we recorded 11 Wordmark requests and 10 or 11 JS requests, that yields 91% and 100% JS support, respectively, but those estimates are unreliable due to a small sample that is most likely not representative of the entire population. A more inclusive choropleth, containing 235 countries, is provided below:

```{r js_support_choropleth_full}
worldwide_js_support <- requests %>%
  group_by(date, country) %>%
  summarize(index_js = sum(index_js), wikipedia_wordmark = sum(wikipedia_wordmark)) %>%
  mutate(proportion = correct_proportion(index_js / wikipedia_wordmark)) %>%
  group_by(country) %>%
  summarize(proportion = median(proportion))

ggplot(data = worldwide_js_support) +
  geom_map(data = world_map, aes(map_id = region), fill = "white", map = world_map, color = "black", size = 0.25) +
  geom_map(aes(map_id = country, fill = proportion), map = world_map) +
  expand_limits(x = world_map$long, y = world_map$lat) +
  scale_fill_gradientn(name = "% of requests with JS support: ",
                       limits = c(0.5, 1),
                       colors = RColorBrewer::brewer.pal(5, "YlGnBu"),
                       labels = scales::percent_format()) +
  ggtitle("JS support (from countries with >100 WP Wordmark requests)") +
  ggthemes::theme_fivethirtyeight() +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(),
        legend.key.width = unit(0.6, "in"))
```

\newpage

## Worldwide Wikipedia Portal Traffic

```{r traffic_choropleth}
requests %>%
  group_by(date, country) %>%
  summarize(wikipedia_wordmark = sum(wikipedia_wordmark)) %>%
  group_by(country) %>%
  summarize(avg = median(wikipedia_wordmark)) %>%
  ggplot(data = .) +
  geom_map(data = world_map, aes(map_id = region), fill = "white", map = world_map, color = "black", size = 0.25) +
  geom_map(aes(map_id = country, fill = avg), map = world_map) +
  expand_limits(x = world_map$long, y = world_map$lat) +
  scale_fill_gradientn(name = "Sampled requests",
                       colors = RColorBrewer::brewer.pal(5, "YlGnBu"),
                       labels = polloi::compress) +
  ggtitle("Sources of sampled requests (median of 4 days)") +
  ggthemes::theme_fivethirtyeight() +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(),
        legend.key.width = unit(0.6, "in"))
```

U.S. accounted for approximately 44% of the total traffic to the portal, with U.K. coming in second at 8%, so it also makes sense to look at a version of this map excluding U.S. traffic.

```{r traffic_choropleth_woUS}
requests %>%
  group_by(date, country) %>%
  summarize(wikipedia_wordmark = sum(wikipedia_wordmark)) %>%
  group_by(country) %>%
  summarize(avg = median(wikipedia_wordmark)) %>%
  ggplot(data = .) +
  geom_map(data = world_map, aes(map_id = region), fill = "white", map = world_map, color = "black", size = 0.25) +
  geom_map(aes(map_id = country, fill = avg), map = world_map) +
  expand_limits(x = world_map$long, y = world_map$lat) +
  scale_fill_gradientn(name = "Sampled requests",
                       limits = c(1e0, 2.3e5),
                       colors = RColorBrewer::brewer.pal(5, "YlGnBu"),
                       labels = polloi::compress) +
  ggtitle("Sources of sampled requests (median of 4 days, excluding U.S.)") +
  ggthemes::theme_fivethirtyeight() +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(),
        legend.key.width = unit(0.6, "in"))
```

\newpage

The following is a table showing the top 20 countries (by % of traffic) and the corresponding approximate percentage of users with JS support.

```{r traffic_top20_table}
requests %>%
  group_by(date, country) %>%
  summarize(wikipedia_wordmark = sum(wikipedia_wordmark), index_js = sum(index_js)) %>%
  group_by(country) %>%
  summarize(`Wordmark DLs` = median(wikipedia_wordmark),
            `JS DLs` = median(index_js)) %>%
  mutate(`% of Total` = sprintf("%.2f%%", 100 * correct_proportion(`Wordmark DLs` / sum(`Wordmark DLs`))),
         `Approx. % with JS support (JS/Wordmark)` = sprintf("%.2f%%", 100 * correct_proportion(`JS DLs` / `Wordmark DLs`))) %>%
  top_n(20, `% of Total`) %>%
  arrange(desc(`Wordmark DLs`)) %>%
  mutate(`Wordmark DLs` = polloi::compress(`Wordmark DLs`),
         `JS DLs` = polloi::compress(`JS DLs`)) %>%
  knitr::kable()
```

```{r traffic_relative_approx_choropleth}
requests %>%
  group_by(date, country) %>%
  summarize(wikipedia_wordmark = sum(wikipedia_wordmark)) %>%
  group_by(country) %>%
  summarize(avg = median(wikipedia_wordmark)) %>% # median across 4 days of data
  mutate(share = avg/sum(avg)) %>%
  ggplot(data = .) +
  geom_map(data = world_map, aes(map_id = region), fill = "white", map = world_map, color = "black", size = 0.25) +
  geom_map(aes(map_id = country, fill = share), map = world_map) +
  expand_limits(x = world_map$long, y = world_map$lat) +
  scale_fill_gradientn(name = "Traffic share",
                       limits = c(0, 0.5),
                       colors = RColorBrewer::brewer.pal(5, "YlGnBu"),
                       labels = scales::percent_format()) +
  ggtitle("Approximate relative traffic") +
  ggthemes::theme_fivethirtyeight() +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(),
        legend.key.width = unit(0.6, "in"))
```

## Browsers and JS

As of 9 Feb 2016, the following browsers account for at least 1% of daily traffic.

```{r browser_setup}
top_browsers <- requests %>%
  group_by(date, browser) %>%
  summarize(wikipedia_wordmark = sum(wikipedia_wordmark)) %>%
  group_by(browser) %>%
  summarize(wikipedia_wordmark = median(wikipedia_wordmark)) %>%
  mutate(share = wikipedia_wordmark/sum(wikipedia_wordmark)) %>%
  keep_where(share > 0.01) %>%
  arrange(desc(share))

browser_js <- requests %>%
  keep_where(browser %in% top_browsers$browser) %>%
  group_by(date, browser) %>%
  summarize(index_js = sum(index_js), wikipedia_wordmark = sum(wikipedia_wordmark)) %>%
  mutate(proportion = index_js / wikipedia_wordmark) %>%
  group_by(browser) %>%
  summarize(`median JS support` = median(proportion)) %>%
  dplyr::left_join(top_browsers, by = "browser")
```

```{r browser_share_table}
top_browsers %>%
  mutate(share = sprintf("%.2f%%", 100 * share),
         wikipedia_wordmark = polloi::compress(wikipedia_wordmark)) %>%
  rename(`Browser` = browser,
         `Median Wordmark Requests Per Day` = wikipedia_wordmark,
         `Median Daily Share` = share) %>%
  knitr::kable(align = c("l", "r", "r"))
```

```{r browser_js_support_lollipopchart}
browser_js %>%
  mutate(`median JS support` = correct_proportion(`median JS support`)) %>%
  ggplot(data = ., aes(y = `median JS support`, x = browser)) +
  geom_pointrange(aes(ymax = `median JS support`, ymin = 0)) +
  geom_point(color = "black", size = 4) +
  geom_point(aes(color = share), size = 3) +
  scale_x_discrete(limits = browser_js$browser[order(browser_js$share, decreasing = FALSE)]) +
  scale_y_continuous(breaks = seq(0, 1.25, 0.10), labels = scales::percent_format()) +
  coord_flip() +
  ggtitle("Proportion of JS support in requests from 'top browsers'") +
  xlab("Browser") +
  ylab("% of Requests With JS Support") +
  scale_color_gradientn(name = "% of Total Traffic",
                       limits = c(0, 0.2),
                       colors = RColorBrewer::brewer.pal(5, "YlGnBu"),
                       labels = scales::percent_format()) +
  ggthemes::theme_gdocs() +
  theme(rect = element_blank())
```

The reader might notice that the proportion of JS support in some of the browsers is around 60% and 80%. We suspect this may be due to users turning off JavaScript in their browser settings for privacy and bandwidth reduction.

\newpage

The next few pages show the popularity of each of the "meta top 10" browsers across the world. To clarify, the "meta top 10" list was compiled by getting the top 10 browsers in each country and then getting the top 10 browsers by number of countries where it was in the top 10.

```{r most_popular_browser_country}
top10_browsers_per_country <- requests %>%
    group_by(date, country, browser) %>%
    summarize(wikipedia_wordmark = sum(wikipedia_wordmark)) %>%
    group_by(country, browser) %>%
    summarize(wikipedia_wordmark = median(wikipedia_wordmark)) %>%
    top_n(10, wikipedia_wordmark)
topN_browsers <- top10_browsers_per_country %>%
  group_by(browser) %>%
  summarize(countries = n()) %>%
  top_n(10, countries) %>%
  { .$browser }
top10_browsers_per_country$browser_major <- ifelse(top10_browsers_per_country$browser %in% topN_browsers,
                                                   top10_browsers_per_country$browser, "Other")
for (i in 1:length(topN_browsers)) {
  p <- top10_browsers_per_country %>% # should now have 1 of browser_major for each country
    dplyr::filter_(~browser_major == topN_browsers[i]) %>%
    ggplot(data = .) +
    geom_map(data = world_map, aes(map_id = region), fill = "white", map = world_map, color = "gray60", size = 0.25) +
    geom_map(aes(map_id = country), fill = "#377eb8", map = world_map, color = "black", size = 0.25) +
    expand_limits(x = world_map$long, y = world_map$lat) +
    ggtitle(paste("Countries where", topN_browsers[i], "is one of top 10 browsers")) +
    ggthemes::theme_fivethirtyeight() +
    theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank())
  print(p)
}; rm(p, i, topN_browsers, top10_browsers_per_country)
```

\newpage

# Appendix

The following Hive query was used to aggregate the request logs:

```
ADD JAR /home/ebernhardson/refinery-hive-0.0.21-SNAPSHOT.jar;
CREATE TEMPORARY FUNCTION is_spider AS 'org.wikimedia.analytics.refinery.hive.IsSpiderUDF';
CREATE TEMPORARY FUNCTION is_wikimedia AS 'org.wikimedia.analytics.refinery.hive.IsWikimediaBotUDF';
CREATE TEMPORARY FUNCTION ua_parser AS 'org.wikimedia.analytics.refinery.hive.UAParserUDF';
USE wmf;
SELECT date, country_code, country, browser, file, COUNT(1) AS requests FROM (
  SELECT
    TO_DATE(ts) AS date,
    geocoded_data['country_code'] AS country_code,
    geocoded_data['country'] AS country,
    CONCAT(ua_parser(user_agent)['browser_family'], ' ',
           ua_parser(user_agent)['browser_major']) AS browser,
    CASE
      WHEN INSTR(uri_path, '/portal/wikipedia.org/assets/js/index-') > 0 THEN 'index_js'
      ELSE 'wikipedia_wordmark' END
    AS file
  FROM webrequest
  WHERE webrequest_source IN('text', 'mobile')
   AND year = 2016 AND month = 2 AND day > 4
   AND uri_host IN('wikipedia.org', 'www.wikipedia.org')
   AND (
     INSTR(uri_path, '/portal/wikipedia.org/assets/js/index-') > 0
     OR
     INSTR(uri_path, '/portal/wikipedia.org/assets/img/Wikipedia_wordmark') > 0
   )
   AND NOT is_spider(user_agent) AND NOT is_wikimedia(user_agent)
   AND NOT (ua_parser(user_agent)['device_family'] = 'Spider')
   AND http_status IN('200', '202', '301', '302', '304')
   AND NOT(geocoded_data['country'] = 'Unknown')
 ) AS requests
 GROUP BY date, country_code, country, browser, file;
```
